---
title: "GWAS"
author: "Juan Jimenez"
date: "05-19-2022"
output: pdf_document
---

## Path Normalization

````{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('')) 
#library(tidyverse)
```

--------------------------------

# Geographic distribution of germplasm


```{r warning=FALSE, eval=T}
### read in the data from step one
df <- read.csv("http://ricediversity.org/data/sets/44kgwas/RiceDiversity.44K.germplasm.csv", skip=1)
df$Latitude <- as.numeric(as.character(df$Latitude))
range(df$Latitude, na.rm = T)
df$Longitude <- as.numeric(as.character(df$Longitude))
range(df$Longitude, na.rm = T)
```


---

# Geographic distribution of germplasm

```{r fig.height=50, fig.width=100, message=TRUE, warning=FALSE, eval=FALSE}
library(ggmap)
##lowerleftlon, lowerleftlat, upperrightlon, upperrightlat
myloc <- c(-105, -40, 170, 56)

mymap <- get_map(location=myloc, source="stamen", crop=FALSE, color="bw")
ggmap(mymap) + 
    geom_point(aes(x = Longitude, y = Latitude), data = df,
               alpha = .9, size = 1, col="red")
```

---

# Geographic distribution of germplasm

### Color code the subpopulatios

```{r fig.height=50, fig.width=100, message=TRUE, warning=FALSE, eval=FALSE}
table(df$Sub.population)
##lowerleftlon, lowerleftlat, upperrightlon, upperrightlat
myloc <- c(-105, -40, 170, 56)

mymap <- get_map(location=myloc, source="stamen", crop=FALSE, color="bw")
ggmap(mymap) + 
    geom_point(aes(x = Longitude, y = Latitude, color=Sub.population), data = df,
               alpha = .9, size = 2)
```

---

# Geographic distribution of germplasm

###  Weight the size of the dots

Count the number of the accessions for each country.

```{r fig.height=50, fig.width=100, message=TRUE, warning=FALSE, eval=FALSE}
library(plyr)
c <- ddply(df, .(Country.of.origin), nrow)
c <- subset(c, Country.of.origin != "")
df2 <- merge(c, df[, c("Country.of.origin",  "Latitude", "Longitude")], by="Country.of.origin")
df2 <- df2[!duplicated(df2$Country.of.origin), ]

mymap <- get_map(location=myloc, source="stamen", crop=FALSE, color="bw")
ggmap(mymap) + 
    geom_point(aes(x = Longitude, y = Latitude, color=Country.of.origin), data = df2,
               alpha = .9, size = df2$V1/3) +
    theme(legend.position = "none")
head(df2[order(df2$V1, decreasing = T),])
```

---

# Genotypic data

- `data/RiceDiversity_44K_Genotypes_PLINK/`
  - sativas413.fam
  - sativas413.map
  - sativas413.ped
- Or download data from: [zipped data](http://ricediversity.org/data/sets/44kgwas/)

--------------------

### PLINK PED File format

__.fam__: A text file with no header line, and one line per sample with the following six fields:

- Family ID ('FID')
- Within-family ID ('IID'; cannot be '0')
- Within-family ID of father ('0' if father isn't in dataset)
- Within-family ID of mother ('0' if mother isn't in dataset)
- Sex code ('1' = male, '2' = female, '0' = unknown)
- Phenotype value ('1' = control, '2' = case, '-9'/'0'/non-numeric = missing data if case/control)

---

# Genotypic data

- `data/RiceDiversity_44K_Genotypes_PLINK/`
  - sativas413.fam
  - sativas413.map
  - sativas413.ped
- Or download data from: [zipped data](http://ricediversity.org/data/sets/44kgwas/)

--------------------

### PLINK PED File format

__.map__: A text file with no header file, and one line per variant with the following 3-4 fields:

- Chromosome code. PLINK 1.9 also permits contig names here, but most older programs do not.
- Variant identifier
- Position in morgans or centimorgans (optional; also safe to use dummy value of '0')
- Base-pair coordinate

---

# Genotypic data

- `data/RiceDiversity_44K_Genotypes_PLINK/`
  - sativas413.fam
  - sativas413.map
  - sativas413.ped
- Or download data from: [zipped data](http://ricediversity.org/data/sets/44kgwas/)

--------------------
### PLINK PED File format

__.ped__: Contains no header line, and one line per sample with 6+2V fields where V is the number of variants.

- The first six fields are the same as those in a `.fam` file. 
- The seventh and eighth fields are allele calls for the first variant in the `.map` file ('0' = no call); 
- the 9th and 10th are allele calls for the second variant; and so on.

---

# Genotypic data manipulation

1. ssh to your HCC account: `ssh USERID@crane.unl.edu`
2. cd to your agro932-lab repo: `cd PATH/courses/2022-agro932-lab`
3. Fetch upstream from the GitHub. (git clone a new copy in a diff path)
4. git pull to sync it: `git pull`
5. request a quick computing node `srun --qos=short --nodes=1 --licenses=common --ntasks=4 --mem 32G --time 6:00:00 --pty bash`

```{bash, eval=FALSE}
mkdir largedata
cp -r data/RiceDiversity_44K_Genotypes_PLINK largedata/

module load plink/1.90
# convert it to binary file
cd largedata/RiceDiversity_44K_Genotypes_PLINK
plink --file sativas413 --make-bed --out binary_sativas413
```

### Calculate MAF and missingness

[plink v1.9](https://www.cog-genomics.org/plink/1.9/)

- `--freq`: writes a minor allele frequency report to plink.frq
- `--missing`: produces sample-based and variant-based missing data reports. 

```{bash, eval=FALSE}
plink -bfile binary_sativas413 --freq --missing --out sativas413

# copy results back to cache folder!
cd ../../
cp largedata/RiceDiversity_44K_Genotypes_PLINK/sativas413.frq cache/
cp largedata/RiceDiversity_44K_Genotypes_PLINK/sativas413.lmiss cache/
```

### Remote cluster

```{bash, eval=FALSE}
git add --all
git commit -m "GWAS frq and missingness"
git push
```

### local computer

`git pull`

---

# Visualize MAF and locus missing rate

```{r, eval=FALSE}
# install.packages("data.table")
library("data.table")

maf <- fread("cache/sativas413.frq", header=TRUE)
lmiss <- fread("cache/sativas413.lmiss", header=TRUE)


pdf("graphs/maf_lmiss.pdf", width = 10, height=5)
par(mfrow=c(1,2))
hist(maf$MAF, breaks=50, col="#cdb79e", main="MAF (SNP = 36,901)", xlab="Minor Allele Freq")
#abline(v=0.01, lty=2, col="black", lwd=3)
abline(v=0.05, lty=2, col="red", lwd=3)

hist(lmiss$F_MISS, breaks=35, col="#cdb79e", main="Missingness (SNP = 36,901)", xlab="Missing Rate")
#abline(v=0.6, lty=2, col="red", lwd=3)
#abline(v=0.05, lty=2, col="red", lwd=3)
dev.off()
```

______________________________________



# LD Decay

Linkage disequilibrium (LD) refers to the nonrandom associations of alleles at different loci. 
- The decay of LD is affected by recombination rate and the number of generations of recombination. Therefore, investigating LD decay may reveal the population recombination history.
- For GWAS, it helps to estimate the number of markers needed.

--

### using PLINK to compute LD

- With `--r2`, when a table format report is requested, pairs with r2 values less than 0.2 are normally filtered out of the report. 

- Use `--ld-window-r2` to adjust this threshold.

---

# LD Decay

- Using HCC to conduct the analysis

```{bash, eval=FALSE}
### log onto HCC
ssh YOUR_USER_ID@crane.unl.edu
# Enter your passcode

### request computing node
srun --qos=short --nodes=1 --licenses=common --ntasks=4 --mem 32G --time 6:00:00 --pty bash
# cd to your project repo
# git pull
```

- Load the `PLINK` module

```{bash, eval=FALSE}
module load plink
cd largedata/RiceDiversity_44K_Genotypes_PLINK/

# By default, when a limited window report is requested, 
# every pair of variants with at least (10-1) variants between them, 
# or more than 1000 kilobases apart, is ignored. 
# You can change the first threshold with `--ld-window`, and the second threshold with `--ld-window-kb`.
plink -bfile binary_sativas413 --r2 --ld-window 100 --ld-window-kb 100 --ld-window-r2 0 --out binary_sativas413 
```

---

# Summarize LD decay rate

```{r, eval=FALSE}
library("data.table")
# cd back to your project home dir
df <- fread("largedata/RiceDiversity_44K_Genotypes_PLINK/binary_sativas413.ld", data.table=FALSE)

BINSIZE = 100
df$dist <- df$BP_B - df$BP_A
df$bin <- round(df$dist/BINSIZE, 0)

library(plyr)

df2 <- ddply(df, .(bin), summarise,
      meanr2 = mean(R2))

write.table(df2, "cache/ld_in_100bp_bin.csv", sep=",", row.names=FALSE, quote=FALSE)
```

---

### Plot LD decay results

- Plot the figure and sync the figure through github

```{r, eval=FALSE}
ld <- read.csv("cache/ld_in_100bp_bin.csv")

pdf("graphs/ld_decay.pdf", width=10, height=10)
plot(ld$bin*100, ld$meanr2, xlab="Physical distance (bp)", ylab="R2", main="LD decay rate in rice")
abline(h=0.3, col="red")
dev.off()

```

--

- Sync the results through github and plot the figure on local computer, git push and git pull in local

```{r, eval=FALSE}
ld <- read.csv("cache/ld_in_100bp_bin.csv")

plot(ld$bin*100, ld$meanr2, xlab="Physical distance (bp)", ylab="R2", main="LD decay rate in rice")
abline(h=0.3, col="red")
```

---

# Population structure using PCA

PCA (principal component analysis) is a method often used to compress the high dimensional data without losing as much information.

Basically, it creates linear combinations of the columns of matrix information, $\mathbf{X}$, and generates, at most, $p$ linear combinations, called principal components.

\begin{align*}
PC_1 = \mathbf{w_1X} \\
PC_2 = \mathbf{w_2X} \\
PC_p = \mathbf{w_pX} \\
\end{align*}
Here, $\mathbf{w_p}$ is the __eigenvector__ of $PC_p$.

The first PC, or PC1, captures the largest variance, the 2nd PC, or PC2, captures the 2nd largest variance, and so on.


---

# PCA using PLINK

- By default, `--pca` extracts the top 20 principal components; you can change the number by passing a numeric parameter. 
- Eigenvectors are written to `plink.eigenvec`, and top eigenvalues are written to `plink.eigenval`. 
- The 'header' modifier adds a header line to the `.eigenvec` file(s).

```{bash, eval=FALSE}
cd largedata/RiceDiversity_44K_Genotypes_PLINK
plink -bfile binary_sativas413 --pca 'header' --out sativas413 
cd ../../
cp largedata/RiceDiversity_44K_Genotypes_PLINK/sativas413.eigenvec cache/
```

---

# PCA using PLINK

### Plot the PCA results in R

```{r, eval=FALSE}
pca <- read.table("cache/sativas413.eigenvec", header=TRUE)
plot(pca$PC1, pca$PC2, xlab="PC1", ylab="PC2")
plot(pca$PC3, pca$PC4, xlab="PC3", ylab="PC4")
```


```{r, eval=FALSE}
#making a plot in R for the PCA
# install.packages("scatterplot3d")
library("scatterplot3d")

fsize=16
pdf("graphs/pca_3d.pdf", width=10, height=10)
scatterplot3d(pca[,3:5], pch = 16, cex.symbol=1.2, color="#00BFC4", main="Maize Diversity Panel", angle=40)
dev.off()
```


---

# GWAS using the `gemma` software package

### Fit the QK model

\begin{align*}
\mathbf{y} &= \mathbf{Qv} + \mathbf{w_i}m_i + \mathbf{Zu} + \mathbf{e} \\
\end{align*}


```{bash, eval=FALSE}
module load gemma
# To calculate centered relatedness matrix (will take ~ 1 min):
gemma -bfile binary_sativas413 -gk 1 -o binary_sativas413
```

--

-9 Phenotype doesn't go well with `gemma`. We have to change a little bit.

---

# GWAS using the `gemma` software package

### Fit the QK model

```{r, eval=FALSE}
#Using R to fit the QK model
library("data.table")

ped <- fread("sativas413.ped", header=FALSE)
ped$V6 <- 1
fwrite(ped, "sativas413.ped", sep="\t", row.names=FALSE, col.names = FALSE, quote=FALSE)

fam <- fread("sativas413.fam", header=FALSE)
fam$V6 <- 1
fwrite(fam, "sativas413.fam", sep="\t", row.names=FALSE, col.names = FALSE, quote=FALSE)
```

--

```{bash, eval=FALSE}
module plink
plink --file sativas413 --make-bed --out binary_sativas413
# To calculate centered relatedness matrix (will take ~ 1 min): Creating the GK matrix
gemma -bfile binary_sativas413 -gk 1 -o binary_sativas413
```


```{r, eval=FALSE}
#Checking the data
library("data.table")
k <- fread("largedata/RiceDiversity_44K_Genotypes_PLINK/output/binary_sativas413.cXX.txt", header=FALSE)
dim(k)
```

---

# GWAS using the gemma software package

### Q matrix
If one has covariates other than the intercept and wants to adjust for those covariates simultaneously, one should provide `GEMMA` with a covariates file containing an intercept term explicitly.
> from Gemma manual

```
1 1 -1.5
1 2 0.3
1 2 0.6
1 1 -0.8
1 1 2.0
```

--

```{r, eval=FALSE}
#Creating the matrix for PCA as covariate
# cd to largedata/RiceDiversity_44K_Genotypes_PLINK
pca <- read.table("sativas413.eigenvec", header=TRUE)
pca[,2] <- 1
write.table(pca[,2:5], "pc3.txt", sep="\t", row.names=FALSE, 
            quote=FALSE, col.names = FALSE)
```

---

# Phenotypic data

```{r fig.height=5, fig.width=5, message=FALSE, warning=FALSE, eval=FALSE}
pheno <- read.delim("http://ricediversity.org/data/sets/44kgwas/RiceDiversity_44K_Phenotypes_34traits_PLINK.txt", header=TRUE)

library(ggplot2)
#Here you can replace x= for the trait you want to visualize in GWAS
ggplot(pheno, aes(x=Plant.height)) + 
  geom_histogram(aes(y=..density..), bins=50, fill="#999999")+
  geom_density(alpha=.2, fill="#FF6666") +
  labs(title="Phenotype histogram plot",x="Plant Height", y = "Density")+
  theme_classic()

```

--

- On HCC, write the `pheno.txt` to the genotypic data folder

```{r, eval=FALSE}
pheno <- read.delim("http://ricediversity.org/data/sets/44kgwas/RiceDiversity_44K_Phenotypes_34traits_PLINK.txt", header=TRUE)
write.table(pheno[, -1:-2], "largedata/RiceDiversity_44K_Genotypes_PLINK/pheno.txt", 
            sep="\t", row.names=FALSE, quote=FALSE, col.names = FALSE)
dim(pheno)
```


---

# GWAS using the gemma software package


\begin{align*}
\mathbf{y} &= \mathbf{Qv} + \mathbf{w_i}m_i + \mathbf{Zu} + \mathbf{e} \\
\end{align*}

```{bash, eval=FALSE}

#here chose -n number of the column you have the trait you want to work with example 13 for panicle.length and -o name of new gwas file panicle.length
gemma -bfile binary_sativas413 -c pc3.txt -k output/binary_sativas413.cXX.txt -p pheno.txt -lmm 4 -n 12 -o Plant.height -miss 0.9 -r2 1 -hwe 0 -maf 0.05

cp output/Plant.height.assoc.txt ../../cache
```

- `lmm`: specify frequentist analysis choice (default 1; valid value 1-4; 1: Wald test; 2:
likelihood ratio test; 3: score test; 4: all 1-3.)
- `n`: specify phenotype column in the phenotype file (default 1); or to specify which
phenotypes are used in the mvLMM analysis
- `o`: specify output file prefix
- `miss`: specify missingness threshold (default 0.05)
- `r2`: specify r-squared threshold (default 0.9999)
- `hwe`: specify HWE test p value threshold (default 0; no test)
- `maf`: specify minor allele frequency threshold (default 0.01)

---

# The Manhattan plot


```{r, eval=FALSE}
library(qqman)
library("data.table")
res <- fread("cache/Plant.height.assoc.txt")

manhattan(x = res, chr = "chr", bp = "ps", p = "p_wald", snp = "rs", col = c("blue4", "orange3"), logp = TRUE)
```


